\appendix
\chapter{Mugda implementation}
$\mugda$ was implemented in the function language Haskell \cite{haskell}, using the Glasgow Haskell compiler (GHC) \cite{jones93glasgow}.

\section{Source file listing}
\begin{itemize}
\item
\texttt{Lexer.x} : the alex lexer file
\item
\texttt{Parser.y} : the happy parser file
\item
\texttt{Concrete.hs} : concrete syntax produced by the parser
\item
\texttt{TraceError.hs} : provides the user a trace when an error occurs
\item
\texttt{ScopeChecker.hs} : turns concrete into abstract syntax
\item
\texttt{Abstract.hs} : produced by the scope-checker, used during type-checking
\item
\texttt{Values.hs} : defines values , evaluation, signature , type-check monad
\item
\texttt{TypeChecker.hs} : type-checking with admissibility
\item
\texttt{Termination.hs} : syntactic termination check
\item
\texttt{Completeness.hs} : size pattern completeness check
\item
\texttt{SPos.hs} : strict positivity checker
\item
\texttt{Main.hs} : the main module
\item
\texttt{example} directory: example input files
\item
\texttt{Makefile} : for compilation
\end{itemize}
\section{Usage}

\mugda as presented was pretty much directly transfered to ASCII syntax:
\begin{itemize}
\item
lists of constructors and clauses are grouped with brackets \texttt{\{} \texttt{\}} and separated with semicolon \texttt{;}
\item
$\EPi{x}{A}{B}$ is written \texttt{(x : A) -> B }
\item
$A \ra B$ is written \texttt{A -> B}
\item
$\ELet{x}{A}{e}{f}$ is written \texttt{let x : A = e in f}
\item
$ \lam{x}e$ is written \verb+\x -> e +
\item
$ \inacc{e}$ is written \texttt{.e} 
\item
$ \infty$ is written \texttt{\#} 
\item
$\s$ is written \texttt{\$}
\item
one line comments are prefixed by \verb+--+
\item
multi-line comments are put between \verb+{-+ and \verb+-}+
\end{itemize}
a \texttt{let} declaration can be prefixed with \texttt{eval}. Then the value will be evaluated after type checking is done.
As an example showing most of the syntactical features, here is the Fibonacci stream example (\texttt{examples/fib.ma}) in text format:
\begin{verbatim}
data Nat : Set {
  zero : Nat;
  succ : Nat -> Nat 
}

fun add : Nat -> Nat -> Nat {
  add zero = \y -> y;
  add (succ x) = \y -> succ (add x y)
}

sized codata Stream : Size -> Set {
  cons : (i : Size) -> Nat -> Stream i -> Stream ($ i)
}
 
fun tail : Stream # -> Stream # {
  tail (cons .# x xs) = xs
}

fun head : Stream # -> Nat {
  head (cons .# x xs) = x
}

fun nth : Nat -> Stream # -> Nat {
  nth zero xs = head xs;
  nth (succ x) xs = nth x (tail xs) 
}

let 1 : Nat = (succ zero)

cofun fib' : (x : Nat ) -> (y : Nat ) 
               -> (i : Size ) -> Stream i {
  fib' x y ($ i) = cons i x (fib' y (add x y) i)
} 

-- fib = 1, 1, 2, 3, 5 , 8 ...
let fib : Stream # = (fib' 1 1 #)

let 4 : Nat = (succ (succ (succ 1)))

-- fib(4) = 5 
eval let fib4 : Nat = nth 4 fib 
\end{verbatim}
Running \texttt{Main examples/fib.ma} yields the console output:
\begin{verbatim}
***** Mugda v1.0 *****
--- scope checking ---
--- type checking ---
--- evaluating ---
fib4 evaluates to (succ (succ (succ (succ (succ zero)))))
\end{verbatim}

\section{Some implementation details}
The alex \cite{alex} and happy \cite{happy} tools were used to generate lexer and parser for $\mugda$.
Most of the Haskell code is monadic, where \emph{monad transformers} \cite{Grabmueller2006MonadTransformers} are used to keep the signature in a state monad, to provide I/O and tracing of errors. The execution of $\mugda$ can be broken into 4 stages:
\paragraph*{Parsing}
The input file is parsed into \emph{concrete syntax}.
\paragraph*{Scope-Checking}
As mentioned in chapter 3, scope-checking is the first step after parsing.
During parsing, it is not known whether an identifier is variable or constructor etc.
Scope-checking produces \emph{abstract syntax} where all identifiers are categorized, if the $\mugda$ program is well-scoped.
Also some syntactic tests like checking linearity of patterns are already done during this stage.
\paragraph*{Type-Checking}
Every declaration is type-checked.
For mutual declarations, the type-checker also checks admissibility and finally invokes the termination-checker. 
\paragraph*{Evaluation}
For all declarations of the form \verb+eval let l : A = e+, \verb+e+ is evaluated and this value is displayed.